Introductory statistics with `intRo`

### 2016-07-19 jcgs-paper-2016/paper.Rmd
Finish up the revision

---
title: "Introductory statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
  `intRo` is a web-based application for performing basic data analysis and statistical routines. Leveraging the power of R and Shiny, `intRo` implements common statistical functions in an extensible modular structure, while including a point-and-click interface for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as Excel and JMP. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: Interactivity, Programming Paradigms, Reactive Programming, Reproducibility, Statistical Software
bibliography: refs.bib
blind: false
---

@@

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even though instructors tend to use a tool like R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms.

@@

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on the design decisions of `intRo`, in particular, its modular and extensible structure. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

This reproducibility framework has three key advantages. First, this eases a student who may be intimidated by programming into the idea that interacting with a user interface is really just a frontend for code. Seeing the correspondence between graphical clicks and printed code will hopefully also lessen the fear of coding that many students may have. Second, an analysis created by `intRo` can be reproduced in an R session to easily assess and extend the results. Finally, "printing" the results of an `intRo` analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility.

>>>>

---
title: "Introductory statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
  `intRo` is a web-based application for performing basic data analysis and statistical routines. Leveraging the power of R and Shiny, `intRo` implements common statistical functions in an extensible modular structure, while including a point-and-click interface for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: Interactivity, Programming Paradigms, Reactive Programming, Reproducibility, Statistical Software
bibliography: refs.bib
blind: false
---

@@

The widespread adoption of R [@r-stat] as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student [@5359977]. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even if the instructor himself/herself uses R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms.

@@

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on the design decisions of `intRo`, in particular, its modular and extensible structure. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

This reproducibility framework has three key advantages. First, this eases a student who may be intimidated by programming into the idea that interacting with a user interface is really just a frontend for code. Seeing the correspondence between graphical clicks and printed code will hopefully also lessen the fear of coding that many students may have. Second, an analysis created by `intRo` can be reproduced in an R session to easily assess and extend the results. Finally, "printing" the results of an `intRo` analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility. These concepts are important because they encourage best practices with regards to disclosure of analysis methods in research [@baggerly2011reproducible; @xie2015].
